Leader functions and qualities across 59 cultures: an exploratory analysis

### 2019-12-03 leadership_across_cultures_contexts.Rmd
Adding ethnographic examples in discussion. Edits in discussion section.

The operationalized qualities and functions (See Table \@ref(tab:qualfuncvars)) were developed from the ethnographic text records in the database. First ZG read a subset of text records and developed initial variable lists and operationalizations of the qualities and functions in those text records. Then, KS coded all 1212 text records (in a random order) for the operationalized qualities, functions, cost, and benefits, while also creating new variables and operationalizations for additional qualities and functions encountered. After coding half of the random sample KS and ZG revised the operational definitions to produce the final list of variables in Table \@ref(tab:qualfuncvars). ZG then followed the same process of coding the entire sample of 1212 text records for the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` operationalized variables. DISCUSS IRR. ZG and KS then discussed every instance of disagreement between their respective coding to produce a consensus matrix, which is used in the analyses here and added to the published data repository (a data frame named *leader_text2*).

@@

This exploratory study extracted the qualities, functions, benefits, and costs of leadership from the ethnographic record. These variables were *not* determined prior to the study but were instead identified and coded during several passes through `r nrow(text_records)` texts on leadership obtained from the eHRAF (see Methods section). Our results therefore represent the broadest and most comprehensive view of the qualities of leaders (i.e., the types of people that occupy leadership roles), the functions of leaders (i.e., what leaders do), and the costs and benefits of leadership for both leaders and followers, across a diverse and culturally representative sample of the ethnographic record. These results reveal universal dimensions of human leadership while also suggesting important variation by group type, subsistence stype, and sex of the leader.

@@

The distribution of text records by group type and subsistence suggests leaders emerge within similar social domains among populations within varied subsistence strategies, but at variable frequencies. See Figure \@ref(fig:grouptype).

@@

Leadership in residential subgroups (e.g., hunting party, work group) was relatively common among hunter-gatherers but relatively rare among agriculturists. Leadership at the surpacommunity level, in contrast, was relatively rare among hunter-gatherers but quite common among agriculturists.

Hunter-gatherers and pastoralists, populations more likely to experience high residential mobility, exhibited greater relative frequencies of leadership within residential groups and kin groups. Pastoralists, however, rely in part on a highly defensible resource base (i.e., livestock) and exhibit leadership beyond the community more so than hunter-gatherers and on par with horticulturalists. Pastoral and horticultural populations often experience elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaborated political leadership systems.

Figure \@ref(fig:grouptype) is consistent with a negative relationships between political leadership at the community leadership or above and leadership in kin groups or residential groups. One possible interpretation is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups erodes. It is interesting, however, that even among the agricultural populations that are characterized by leadership systems beyond the community level, leadership within residential and family group persists.

@@

Across a broad range of cultures and social contexts leaders are nearly universally high status males who are described as knowledgeable or intelligent and who are experienced and have a demonstrated history of previous accomplishments or special skills. They most frequently resolve conflicts within their respective groups while also providing important social functions, organizing cooperation among followers, providing counsel or input in followers' affairs, and distributing resources. See Figure \@ref(fig:varsupport). These qualities and functions correspond to those highlighted by several evolutionary theories of leadership [@garfield_review_2019]. See the *Synthesis* section below.

@@

Several functions and qualities that are theoretically important were relatively rare in the ethnographic record. In the literature on leadership among non-human animals, for example, leaders commonly initiate group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions that @smith_leadership_2016 identified as comparable among humans and non-human animals, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures. Several theories of leadership also emphasize the moral authority of leaders [e.g., @henrich_big_2015; @henrich_evolution_2001; @richerson_tribal_2009; @tietjen_moral_1985; @boggild_evolved_2016], yet leaders were seen as sources of moral authority or "fair" in only a minority of cultures (`r moral_pct`% and `r fairness_pct`% of cultures, respectively). The "Dominance" theory of leadership [e.g., @chapais_competence_2015; @cheng_dominance_2019; @cheng_pride_2010; @henrich_evolution_2001; @tiger_imperial_1971] emphasizes leaders' ability to instill fear, but in only `r feared_pct`% of cultures were leaders characterized as feared. Finally, a large literature on leadership in industrialized societies highlights the attractiveness of leaders [e.g., @sinclair_sexuality_1995; @altemeyer_sexual_1974; @march_leadership_2009], but leaders were characterized as attractive in only `r attractive_pct`% of cultures.

@@

Leadership and high social status are expected to be associated with particular suites of characteristics, which potentially vary by cultural or social context, personality, or via other inter-individual variation [@cheng_two_2013; @von_rueden_roots_2014; @hogan_personality_2020; @lukaszewski_extraversion_2015; @henrich_evolution_2001]. Exploratory cluster analyses reported here suggest multiple leadership phenotypes and associated features.

We performed hierarchical cluster analysis on our leader quality and leader function variables to determine if these variables formed theoretically meaningful categories. Variables were clustered together if they were positively correlated. Leader qualities formed two top-level clusters, which we identify as a *Material, social, and reproductive success* cluster, and a *Social orientation* cluster. See Figure \@ref(fig:cluster)A.

Most of the leader quality variables were found in the *Social orientation* cluster, which had two smaller and more weakly supported subclusters that we identified as *Ability to provide social benefits* and *Ability to impose social costs*. The social benefit subcluster contained the *Knowledgeable/intelligent* and *Experienced/accomplished* variables that have been emphasized by the *Prestige* theory [@henrich_evolution_2001; @cheng_two_2013] and the more general *Computational services* theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and these variables are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport), a point to which we shall return. The remaining social benefit variables included a mix of positive personal traits, such as generosity and proper behavior, as well as interpersonal skills, having a favorable personality, oratory skill, and charisma, which correspond to a charismatic style of leadership that has been identified in industrialized settings [@den_hartog_culture_1999; @grabo_charismatic_2016; @howell_role_2005]. A few variables did not seem to fit the social benefit theme, however, such as *Xenophobia* and *Drug consumption* (although drug use is often an important social activity).

@@

Leaders, probably as a consequence of their ability to provide social benefits and impose social costs, are also characterized by their material, social, and reproductive success (top cluster, Figure \@ref(fig:cluster)A), which is consistent with much of the theoretical literature [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013]. Again, some variables in this subcluster, such as *Ingroup member* and *Loyalty*, did not quite fit the theme.

@@

The leader function variables formed two major clusters, one that we identified as "Prosociality" and one that we identified as "Management". The prosocial cluster included *Distribute resources*, *Protection*, and *Prosocial investment*, which are functions that are commonly highlighted in the anthropological literature on leadership [@sahlins_poor_1963; @johnson_organizational_1982; @service_origins_1975; @boehm_egalitarian_1993; @boehm_hierarchy_1999; @johnson_evolution_1987].

The management subcluster included many more variables, which were grouped into two subclusters that we identified as "Mediation" and "Organization". The mediation subcluster included variables such as *Resolve conflict* (the most widely supported function in our sample) and *Provide counsel* (also widely supported), which again are functions that are commonly noted in the anthropological literature [@wiessner_risk_1982; @wiessner_collective_2019; @boehm_hierarchy_1999; @knauft_violence_1991]. Other variables in this subcluster indicated more generic group strategizing, such as *Strategic planning* and *Group representative* [@xx; @yy].

The larger organization subcluster included variables such as *Organize cooperation*, *Ritual functions*, and *Military command*. This subcluster aligns with evolutionary theories of leadership that emphasize the role of leaders in organizing collective actions [@hooper_theory_2010; @van_vugt_cognitive_2007; for review, see @garfield_evolutionary_2019].

@@

The evolution of leadership and followership is somewhat perplexing. Leadership often involves the provisioning of a public good at an individual cost, raising the question of how the individual benefits of leadership outweigh the costs. Followership, on the other hand, involves relinquishing individual decision-making to a leader who might or might not act in the follower's interest. The benefits of followership must also outweigh the costs. We therefore coded all benefits and costs of both leadership and followership. Leaders were widely reported to receive material benefits and social status, but loss of material resources and social status were also widely reported. The results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Our results cannot determine, however, if the benefits to leaders and followers outweighed their costs. See Figure \@ref(fig:varsupportcb).

@@

In general, our results conform to the characteristics of leaders that are emphasized in the literature on the evolution of leadership [for review, see @garfield_evolutionary_2019]. Leaders in the ethnographic record are portrayed as generally prosocial, and serve to resolve conflicts and organize cooperation. Leaders also receive material, social, and reproductive benefits. There is also a "dominant" style of leadership involving physical formidability, fear, and aggression.

One interesting pattern that has not been widely discussed in the evolutionary literature on leadership is the role of shamans and their intersection, in our results, with the dominant style of leadership. Our social cost cluster (Figure \@ref(fig:cluster)A) includes the *Supernatural* abilities variable. @winkelman1987cross found that in 47% of societies sampled from the HRAF ($n=43$) shamans were associated with providing decision-making services and associated with military power. Providing ritual functions and military leadership are common (documented in over 50% of societies here) functions leaders serve (see Figure \@ref(fig:varsupport)). Widespread preferences for egalitarianism and social equality are likely strong drivers of leader preferences across diverse societies [@cheng_dominance_2019], but shamanism and ritual based contexts represent a social domain in which ideologies support dominance-based leadership.

We also found a link between prestige-based leadership, associated with deference towards individuals with special skills, high intelligence, and accomplishments [@cheng_dominance_2019; @van_vugt_dual_2019; @henrich_evolution_2001], and charismatic leadership, associated with oratory skills, high charisma, confidence, and interpersonal skills [@vanvugtXXX]. See, Figure \@ref(fig:cluster)A, *Ability to provide social benefits* cluster. This suggests that these leadership strategies might be closely related, or part of a broader strategy of social influence based on the provisioning of various benefits.

@@

@@

Contemporary theoretical models of the evolution of leadership often emphasize leaders roles in organizing collective action and in managing between group conflicts [@hooper_theory_2010; @pietraszewski_evolution_2019; @price_evolution_2014]. Results here do support the important functions of leaders in organizing cooperation and facilitating social functions -- services which provide direct benefits to the group. However, the ubiquity of conflict resolution as a function of leaders across social contexts calls for greater attention to the roles of leaders of all types in managing intra-group conflict between group members -- services tailored to provide direct benefits to a few individuals and more indirect group benefits. [@glowacki_leadership_2015; @wiessner_collective_2019; @pupu_challenges_2018].

Conflict resolution, decision-making, providing counsel. Dominance. Benefits/costs.

>>>>

The operationalized qualities and functions (See Table \@ref(tab:qualfuncvars)) were developed from the ethnographic text records in the database. First ZG read a subset of text records and developed initial variable lists and operationalizations of the qualities and functions in those text records. Then, KS coded all 1212 text records (in a random order) for the operationalized qualities, functions, cost, and benefits, while also creating new variables and operationalizations for additional qualities and functions encountered. After coding half of the random sample KS and ZG revised the operational definitions to produce the final list of variables in Table \@ref(tab:qualfuncvars). ZG then followed the same process of coding the entire sample of 1212 text records for the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` operationalized variables. ZG and KS then discussed every instance of disagreement between their respective coding to produce a consensus matrix, which is used in the analyses here and added to the published data repository (a data frame named *leader_text2*).

<!-- DISCUSS IRR. -->

@@

This exploratory study extracted the qualities, functions, benefits, and costs of leadership from the ethnographic record. These variables were *not* determined prior to the study but were instead identified and coded during several passes through `r nrow(text_records)` texts on leadership obtained from the eHRAF (see Methods section). Our results therefore represent the broadest and most comprehensive view of the qualities of leaders (i.e., the types of people that occupy leadership roles), the functions of leaders (i.e., what leaders do), and the costs and benefits of leadership for both leaders and followers across a diverse and culturally representative sample of the ethnographic record. These results reveal universal dimensions of human leadership while also suggesting important variation by group type, subsistence stype, and sex of the leader.

@@

The distribution of text records by group and subsistence type suggests leaders emerge within similar social domains among populations with varied subsistence strategies, but at variable frequencies. See Figure \@ref(fig:grouptype).

@@

Leadership in residential subgroups (e.g., hunting party, work group) was relatively common among hunter-gatherers but relatively rare among agriculturists. Leadership at the surpacommunity level, in contrast, was relatively rare among hunter-gatherers but quite common among agriculturists. The only example of supracommunity leadership among hunter-gatherers was from the Adamans in LOCATION, by AUTHOR [@az02-001] (45). (reproducing AUTHOR), who discussed the roles of a "head chief" from a permnant residential location and their marginal influence over other, local residential leaders:

>Their domestic policy may be described as a communism modified by the authority, more or less nominal, of the chief. The head chief of a tribe is called maia igla and the elders, or sub-chiefs, i.e. those in authority over each community, consisting of from 20 to 50 individuals, maicla . The head chief, who usually resides at a permanent encampment, has authority over all the sub-chiefs, but his power, like theirs, is very limited. It is exercised mainly in organising meetings between the various communities belonging to his tribe, and in exerting influence in all questions affecting the welfare of his followers. It is the chief alone, as may be supposed, who directs the movements of a party while on hunting and fishing expeditions, or when migrating. It is usually through his intervention that disputes are settled, but he possesses no power to punish or enforce obedience to his wishes, it being left to all alike to take the law into their own hands when aggrieved. The aryoto and eremtaga in each tribe have their own head chief, who are independent the one of the other."

Hunter-gatherers and pastoralists, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013], exhibited greater relative frequencies of leadership within residential groups and kin groups. Pastoralists, however, rely in part on a highly defensible resource base (i.e., livestock) and exhibit leadership beyond the community more so than hunter-gatherers and on par with horticulturalists. Pastoral and horticultural populations often experience elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaborated political leadership systems [@ember_warfare_1992]. For example, among the Somali, a politically complex and socially stratigied agro-pastoralists group, DESCIRBE, AUTHOR [@mo04-021] (p. 58) notes that, "the Sultan does not lead the army of the tribe in war. He names a chief of the armed forces."

Figure \@ref(fig:grouptype) is consistent with a negative relationships between political leadership at the community leadership or above and leadership in kin groups or residential groups. One possible interpretation is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes It is interesting, however, that even among the agricultural populations that are characterized by leadership systems beyond the community level, leadership within residential and family group persists. For example, among rural Korean populations a chairwomen of local organizations maintain imporant leadership positions within the groups na drepresent the group to govermental forces. AUTHOR [@aa01-062] (p. 173) describes that in these local 'clubs', "the procedure of discussion and decision-making is not based on an idea of democracy but rather on that of unanimity. In this context, the ‘leader’ does not have to be a person of marked influence in the decision-making process. She is not a figurehead but rather a middleman in the meeting, functioning, in effect, as a go-between for the group. There is no despotic idea in the context of a total consensus in this essentially egalitarian society. The chairwoman stands as a middleman between the government and the village. Although the government may call her a ‘leader’ in the village, she properly does not occupy this role."

@@

Across a broad range of cultures and social contexts leaders are nearly universally high status males who are described as knowledgeable or intelligent and who are experienced and have a demonstrated history of previous accomplishments or special skills. They most frequently resolve conflicts within their respective groups while also providing important social functions, organizing cooperation among followers, providing counsel or input in followers' affairs, and distributing resources. AUTHOR [@oc06-034] (p. 236) in describing kin group leaders among the Iban in LOCATION provides example of this suite of traits in leaders:

>"Bravery may be considered as a means of acquiring status in what Freeman has identified as an egalitarian society. The leader of a longhouse, the tuai rumah, is elected by both men and women of the longhouse for his skills in oratory, warfare, and farming, and his familiarity with adat , or customary law, in order to resolve disputes (Freeman 1970:111). Although usually a man, women tuai rumah are not unknown. The tuai rumah must inspire the confidence of his followers and exemplify the qualities of bravery, endurance, and determination that are essential for survival. Skill in farming and warfare is important in the pioneering context, in which the tuai rumah leads a group to establish a new settlement. The tuai rumah has no authority to command others, but is empowered to resolve disputes through influence and by reference to custom."

See Figure \@ref(fig:varsupport). These qualities and functions correspond to those highlighted by several evolutionary theories of leadership which have drawn heavily on select ethnographic cases [@garfield_review_2019]. See the *Synthesis* section below.

@@

Several functions and qualities that are theoretically important were relatively rare in the ethnographic record. In the literature on leadership among non-human animals, for example, leaders commonly initiate group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions that @smith_leadership_2016 identified as comparable among humans and non-human animals, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures, of which the vast majority (71%) were hunter-gatherers. Among the Blackfoot in the North American plains for example, AUTHOR [@nf06-061] (p. 411) descirbeds that, "If there was a dispute as to whether the camp should move north or south, the chief might present his arguments, gain support, and then say that he was going south. He did not order the others to follow, but he knew that they probably would go with him."

Several theories of leadership also emphasize the moral authority of leaders [e.g., @henrich_big_2015; @henrich_evolution_2001; @richerson_tribal_2009; @tietjen_moral_1985; @boggild_evolved_2016], yet leaders were seen as sources of moral authority or "fair" in only a minority of cultures (`r moral_pct`% and `r fairness_pct`% of cultures, respectively). The *Dominance* theory of leadership [e.g., @chapais_competence_2015; @cheng_dominance_2019; @cheng_pride_2010; @henrich_evolution_2001; @tiger_imperial_1971] emphasizes leaders' ability to instill fear, but in only `r feared_pct`% of cultures were leaders characterized as feared. Finally, a large literature on leadership in industrialized societies highlights the attractiveness of leaders [e.g., @sinclair_sexuality_1995; @altemeyer_sexual_1974; @march_leadership_2009], but leaders were characterized as attractive in only `r attractive_pct`% of cultures.

@@

Leadership and high social status are expected to be associated with particular suites of characteristics, which potentially vary by cultural or social context, personality, or via other inter-individual variation [@cheng_two_2013; @von_rueden_roots_2014; @hogan_personality_2020; @lukaszewski_extraversion_2015; @henrich_evolution_2001]. Exploratory cluster analyses reported here suggest multiple leadership phenotypes.

We performed hierarchical cluster analysis on our leader quality and leader function variables to determine if these variables formed theoretically meaningful categories. Variables were clustered together if they were positively correlated. Leader qualities formed two top-level clusters, which we identify as a *Material, social, and reproductive success* cluster, and a *Social orientation* cluster. See Figure \@ref(fig:cluster)A. Several of traits often cooccur. Among the Trobriands horticulturalists in LOCATION, AUTHOR [@ol06-037] (p. 47 ) describes relationships between plogynous marriages, increased wealthm, and widespread social contacts:

>Chiefs managed unusually large quantities of resources and it is agreed that the basis of such wealth was polygamy. Rank was quite directly expressed in the number of wives. The chief of Omarakana had 16 wives in Malinowski's time, while his predecessor Enamakala had 19 in 1893 (Annual Report 1893-4:19). In 1950 despite long-established missionary activity, there were still 13 (Powell 1960: 136). In traditional Trobriand society it seems that all leaders of village clusters of guyau status became polygamists. An emerging leader might also expect to be given wives by traditional allies outside the cluster while he might demand them from traditional rivals of subordinate subclans (Powell 1960:136)...

Most of the leader quality variables were found in the *Social orientation* cluster, which had two smaller and more weakly supported subclusters that we identified as *Ability to provide social benefits* and *Ability to impose social costs*. The social benefits subcluster contained the *Knowledgeable/intelligent* and *Experienced/accomplished* variables that have been emphasized by the *Prestige* theory [@henrich_evolution_2001; @cheng_two_2013] and the more general *Computational services* theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and these variables are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport), a point to which we shall return. As brief examples of the very many in the ethngorraphic record, among the Azande, "though the most knowledgeable witch-doctors tend to become leaders, there is no institutionalized organization or leadership among them." [fo07-056] (p. 85); and among the Highland Scots, AUTHOR [es10-003] (p. 147) notes, "on the whole, their long ties with place and people, coupled with education, made them useful leaders of a community in which leaders were needed to adapt life to a changing economy." Such exceptionally talented leaders, however, also face unique challenges, particularly within more egalitarian societies which emphasizes personal autonomy and resit heirachy in the context of cultural transitions. AUTHOR [@nm09-056] (p. 183) describes Minnie Kellog (born in 1880), a land-claims activist, writer, and perhaps the most famous female leader among the Iriquous:

>Despite her exceptional gifts -- a brilliant mind, beauty, selfconfidence, unusual oratorical abilities, and her educational attainments -- Kellogg is also the most controversial Iroquois leader of the twentieth century. It is clear from her many bizarre involvements that she misused her prodigious talents and/or was incapable of carrying out all the massive designs she had for her people's betterment. Although acknowledged today as a major force and brilliant person, she is accused by many Iroquois elders of swindling them out of hundreds of thousands of dollars in her abortive efforts to bring their land claims to fruition and of creating debilitating factionalism that impeded tribal development for decades. Unfortunately, because of her questionable ethies and her inability to carry out what she espoused, Kellogg is blamed today for all that went wrong in Iroquois history in the interwar period. Consequently, her life story had the feel of a Greek tragedy: she wanted to use her extraordinary abilities to help her people but ended up accused by them of being a common outlaw.

The remaining social benefit variables included a mix of positive personal traits, such as generosity and proper behavior, as well as interpersonal skills, having a favorable personality, oratory skill, and charisma, which correspond to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @howell_role_2005]. A few variables did not seem to fit the social benefit theme, however, such as *Xenophobia* and *Drug consumption* (although drug use is often an important social activity).

@@

Leaders, probably as a consequence of their ability to provide social benefits and impose social costs, are also characterized by their material, social, and reproductive success (top cluster, Figure \@ref(fig:cluster)A), which is consistent with much of the theoretical literature focused on associated outcomes of leadership and social hierarchy [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013]. Again, some variables in this subcluster, such as *Ingroup member* and *Loyalty*, did not quite fit the theme.

@@

The leader function variables formed two major clusters, one that we identified as "Prosociality" and one that we identified as "Management". The prosocial cluster included *Distribute resources*, *Protection*, and *Prosocial investment*, which are functions that are commonly highlighted in the anthropological literature on leadership in nonindustrial societies [@sahlins_poor_1963; @johnson_organizational_1982; @service_origins_1975; @boehm_egalitarian_1993; @boehm_hierarchy_1999; @johnson_evolution_1987]. In the absenece of institutionalized leadership roles, which charachterize many of the relatively small-scale populations in this sample, leaders often maintain social influence via continued community support and the demands of followers can shape the functions of leaders to meet local needs.

The management subcluster included many more variables, which were grouped into two subclusters that we identified as "Mediation" and "Organization". The mediation subcluster included variables such as *Resolve conflict* (the most widely supported function in our sample) and *Provide counsel* (also widely supported), which again are functions that are commonly noted in the anthropological literature [@wiessner_risk_1982; @wiessner_collective_2019; @boehm_hierarchy_1999; @knauft_violence_1991]. Other variables in this subcluster indicated more generic group strategizing, such as *Strategic planning* and *Group representative* [@de_cremer_intergroup_2002]. In the absenece of state-instiutions local leaders are often tasked with managing various economic systems and providing social services.

The larger organization subcluster included variables such as *Organize cooperation*, *Ritual functions*, and *Military command*. This subcluster aligns with evolutionary theories of leadership that emphasize the role of leaders in organizing collective actions [@hooper_theory_2010; @van_vugt_cognitive_2007; @glowacki_formation_2016; for review, see @garfield_evolutionary_2019].

@@

The evolution of leadership and followership is somewhat perplexing. Leadership often involves the provisioning of a public good at an individual cost, raising the question of how the individual benefits of leadership outweigh the costs. Followership, on the other hand, involves relinquishing individual decision-making to a leader who might or might not act in the follower's interest. The benefits of followership must also outweigh the costs. We therefore coded all benefits and costs of both leadership and followership to better understand which benefits and costs are most commonly associated with leading or following. Leaders were widely reported to receive material benefits and social status, but loss of material resources and social status were also widely reported. The results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Our results cannot determine, however, if the benefits to leaders and followers outweighed their costs, but shed light on the social and biological pressures associated with leadership and social hierarchy. See Figure \@ref(fig:varsupportcb).

@@

In general, our results conform to the characteristics of leaders that are emphasized in the literature on the evolution of leadership and provide broad cross-cultural support for several influential theoretical predictions [for review, see @garfield_evolutionary_2019]. Leaders in the ethnographic record are portrayed as generally prosocial, and serve to resolve conflicts and organize cooperation. Leaders also receive material, social, and reproductive benefits. There is also a "dominant" style of leadership involving physical formidability, fear, and aggression.

One interesting pattern that has not been widely discussed in the evolutionary literature on leadership is the role of shamans and their intersection, in our results, with the dominant style of leadership. Our social costs cluster (Figure \@ref(fig:cluster)A) includes the *Supernatural* abilities variable. @winkelman1987cross found that in 47% of societies sampled from the HRAF ($n=43$) shamans were associated with providing decision-making services and associated with military power. Providing ritual functions and military leadership are common (documented in over 50% of societies here) functions leaders serve (see Figure \@ref(fig:varsupport)). Widespread preferences for egalitarianism and social equality are likely strong drivers of leader preferences across diverse societies [@cheng_dominance_2019], but shamanism and ritual based contexts represent a social domain in which ideologies support dominance-based leadership.

We also found a link between prestige-based leadership, which is associated with deference towards individuals with special skills, high intelligence, and accomplishments [@cheng_dominance_2019; @van_vugt_dual_2019; @henrich_evolution_2001], and charismatic leadership, which is associated with oratory skills, high charisma, confidence, and interpersonal skills [@grabo_charismatic_2016; @howell_role_2005]. See, Figure \@ref(fig:cluster)A, *Ability to provide social benefits* cluster. This suggests that these leadership strategies might be closely related across a broad range of societies and social contexts, or part of a broader strategy of social influence based on the provisioning of various benefits.

@@

The strong support for material and social benefits assocaited with leadership (Figure \@ref(fig:varsupportcb)) and the colinearity of measures associated with material wealth, social capital, and reproductive success (Figure \@ref(fig:cluster)A), suggests processes of leadership and followership might be strong drivers of the emergence and intergenerational transmission of inequlity [@bowles_emergence_2010; @mattison_evolution_2016]. Various sources of material wealth and strategic or instituionalized processes of nepotism are widespread cross-culturally, including among many societies often described as "egalitarian." Theses results suggests anthropologists and economists investigating the emergence and evolution of inequality, including material, social, and reproductive dimensions, should incorporates and pay close attention to group processes and social strategies associated with leadership and followership.

@@

Contemporary theoretical models of the evolution of leadership often emphasize leaders roles in organizing collective action and in managing between group conflicts [@hooper_theory_2010; @pietraszewski_evolution_2019; @price_evolution_2014]. Results here do support the important functions of leaders in organizing cooperation and facilitating social functions -- services which provide direct benefits to the group. However, the ubiquity of conflict resolution by leaders across social contexts and cultures calls for greater attention to the roles of leaders of all types in managing intra-group conflict between group members -- services tailored to provide direct benefits to a few individuals and more indirect group benefits [@glowacki_leadership_2015; @wiessner_collective_2019; @pupu_challenges_2018].

decision-making, providing counsel. Dominance. Benefits/costs.
